Text-mining clinically relevant cancer biomarkers for curation into the CIViC database

### 2018-12-03 paper/paper.Rmd
Substantial paper edits

## Abstract

Precision oncology involves analysis of individual cancer samples to understand the genes and pathways involved in the development and progression of a cancer. To improve patient care, knowledge of diagnostic, prognostic, predisposing and drug response markers is essential. Several knowledgebases have been created by different groups to collate current knowledge on these efforts. These include the open-access Clinical Interpretations of Variants in Cancer (CIViC) knowledgebase. These databases rely on time-consuming manual curation from highly-skilled experts who read and interpret the relevant biomedical literature. To aid in this curation and provide the greatest coverage for these databases, particularly CIViC, we propose the use of text mining approaches to extract these clinically relevant markers from all available published literature. A group of cancer genomics experts annotated biomarkers discussed in 800 sentences and achieved good inter-annotator agreement. We used a supervised learning approach to construct the CIViCmine knowledgebase and extracted `r paper.sentenceCount` relevant sentences from PubMed abstracts and Pubmed Central Open Access full text papers.  CIViCmine contains over `r paper.biomarkerCount` biomarkers associated with `r paper.geneCount` genes, `r paper.drugCount` drugs and `r paper.cancerCount` cancer types, representing `r paper.abstractCount` abstracts and `r paper.articleCount` full-text publications. Through integration with CIVIC, we provide a prioritised list of curatable biomarkers as well as a resource that is valuable to other knowledgebases and precision cancer analysts in general.

@@

The ability to stratify patients into groups that are clinically related is an important step towards a personalized approach to cancer. Over time, a growing number of biomarkers have been developed in order to select patients who are more likely to respond to certain treatments. These biomarkers have also been valuable for prognostic purposes and for understanding the underlying disease biology by defining different molecular subtypes of cancers that should be treated in different ways (e.g. *ERBB2*/*ER*/*PR* testing in breast cancer [@onitilo2009breast]). Immunohistochemistry techniques are the primary approach for testing samples for diagnostic markers. (e.g. *CD15* and *CD30* for Hodgkin’s disease [@rudiger1998differential]). Recently, the lower cost and increasing speed of sequencing has allowed the DNA and RNA of individual patient samples to be characterized for clinical applications [@prasad2016precision]. Throughout the world, this technology is beginning to inform clinician decisions on which treatments to use [@shrager2014rapid]. Such efforts are dependent on comprehensive and current understanding of the clinical relevance of variants. For example, the Personalized Oncogenomics project at the BC Cancer Agency identifies somatic events in the genome such as point mutations, copy number variations and large structural changes and, in conjunction with gene expression data, generates a clinical report to provide an ‘omic picture of a patient’s tumor [@jones2010evolution].

The huge genomic variability in cancers means that each patient sample includes a huge number of new mutations, many of which have never been documented before [@chang2016identifying]. The phenotypic impact of most of these mutations is difficult to discern. This problem is exacerbated by the driver/passenger mutation paradigm where only a fraction of mutations are essential to the cancer (drivers) while many others have occurred through mutational processes that are irrelevant to the cancer and are deemed to have simply come along for the ride (passengers). An analyst trying to understand a new patient sample typically performs a literature review for each gene and specific variant. This is needed to understand its relevance in a cancer type, characterize the driver/passenger role of its observed mutations, and gauge the relevance for clinical decision making.

Several groups have built their own in-house knowledge bases which are developed as analysts examine increasing numbers of cancer patient samples. This tedious and largely redundant effort represents a substantial interpretation bottleneck impeding the progress of precision medicine [@good2014organizing].  To encourage a collaborative effort, the CIViC database (https://civicdb.org) was launched to provide a wiki-like editable online resource where edits and additions are moderated by experts in order to maintain high quality [@griffith2017civic]. The resource provides information about clinically-relevant variants in cancer. Variants include protein-coding point mutations, copy number variations, epigenetic marks, gene fusions, aberrant expression levels and other ‘omic events. It supports four types of biomarkers (also known as evidence types).

Diagnostic evidence items describe variants that can help a clinician diagnose or exclude a cancer. For instance, the JAK2 V617F mutation is a major diagnostic criterion for myeloproliferative neoplasms to identify polycythemia vera, essential thrombocythemia and primary myelofibrosis. Predictive evidence items describe variants that help predict drug sensitivity or response and are valuable in deciding further treatments. Predictive evidence items often explain mechanisms of resistance in patients who progressed on a drug treatment. For example, the ABL1 T315I missense mutation in the BCR-ABL fusion, predicts poor response to imatinib, a tyrosine kinase inhibitor that would otherwise effectively target BCR-ABL, in patients with chronic myeloid leukemia. Predisposing evidence items describe germline variants that increase the likelihood of developing a particular cancer, such as BRCA1 mutations for breast/ovarian cancer or RB1 mutations for retinoblastoma. Lastly, prognostic evidence items describe variants that predict survival outcome. As an example, colorectal cancers that harbor a KRAS mutation are predicted to have worse survival.

@@

Several other groups have created knowledge bases to aid clinical interpretation of cancer genomes. Many of these projects have joined the Variant Interpretation for Cancer Consortium (VICC, http://cancervariants.org/) to coordinate these efforts and have created a federated search mechanism to allow easier analysis across multiple knowledge bases [@wagner2018harmonized]. The CIViC project is co-leading this effort along with OncoKB [@chakravarty2017oncokb], the Cancer Genome Interpreter [@tamborero2018cancer], Precision Medicine Knowledge base [@huang2017cancer], Molecular Match, JAX-Clinical Knowledge base [@patterson2016clinical] and others.

@@

The literature on clinically relevant cancer mutations is growing at an extraordinary rate. For instance there were only 5 publications with BRAF V600E in title or abstract in PubMed in 2004 compared to 454 citations in 2017. In order to maintain a high quality and up-to-date knowledge base, a curation pipeline must be established. This typically involves a queue for papers, triaging those that should be curated and then assignment to a highly experienced curator. This prioritisation step is immensely important given the limited time of curators and the potentially vast number of papers to be reviewed. Prioritisation must identify papers that contain knowledge that is of current relevance to users of the knowledge base. For instance, selecting papers for drugs that are no longer clinically approved would not be valuable to the knowledge base.

Text mining methods have become a common approach to help prioritise papers. These methods fall broadly into two main categories, information retrieval (IR) and information extraction (IE). IR methods focus on paper-level information and can take multiple forms. Complex search queries for specific terms or paper metadata (helped by the MeSH term annotations of papers in biomedicine) are common tools for curators. More advanced document clustering and topic modelling systems can use semi-supervised methods to predict whether a paper would be relevant for curation. Examples of this approach include the document clustering method used for the ORegAnno project [@aerts2008text].

IE methods extract structured knowledge directly from the papers. This can take the form of entity recognition, by explicitly tagging mentions of biomedical concepts such as genes, drugs and diseases. A further step can involve relation extraction to understand the relationship discussed between tagged biomedical entities. This structured information can then be used to identify papers relevant for the knowledge base. IE methods are also used for automated knowledge base population without a manual curation step. For example, the mirTex knowledge base, which collates microRNA and their targets, uses automated relation extraction methods to populate the knowledge base [@li2015mirtex]. Protein-protein interaction networks (such as STRING [@szklarczyk2016string]) are often built using automatically generated knowledge bases.

The main objective of this project was to identify frequently discussed cancer biomarkers which fit the CIViC model but are not yet included in the CIViC knowledge base. We developed an IE-based method to extract key parts of the evidence item: cancer type, gene, drug (where applicable) and the specific evidence type from published literature. This allows us to count the number of mentions of specific evidence items in abstracts and full text articles and compare against the CIViC knowledge base. This paper will present our methods to develop this resource, known as CIViCmine (http://bionlp.bcgsc.ca/civicmine/). This main contributions of this work are an approach for knowledge base construction that could be applied to many areas of biology and medicine, a machine learning method for extracting complicated relationships between four entity types, and extraction of relationships across the largest possible publically accessible set of abstracts and full text articles. This resource, containing `r paper.biomarkerCount` biomarkers, is valuable to all cancer knowledge bases to aid their curation and also as a tool for precision cancer analysts searching for biomarkers not yet included in any other resource.

@@

The full PubMed and PubMed Central Open Access subset corpora was downloaded from the NCBI FTP website using the PubRunner infrastructure [@anekalla2017pubrunner]. These documents were converted to the BioC format for processing with the Kindred package [@lever2017painless]. HTML tags were stripped out and HTML special characters converted to Unicode. Metadata about the papers were retained including PubMed IDs, titles, journal information and publication date. Subsections of the paper were extracted using a customised set of acceptable section headers such as “Introduction”, “Methods”, “Results” and many synonyms of these. The corpora were downloaded in bulk in order to not overload the EUtils RESTFUL service that is offered by the NCBI. In order to avoid duplications of publications in PMCOA and PubMed, the PMIDs of all documents included in PMCOA were used to filter out abstracts from the PubMed corpus. The update files from PubMed were also processed to identify the latest version of each abstract to process.

@@

Term lists were curated for genes, diseases and drugs based on several resources. The cancer list was curated from a section of the Disease Ontology [@schriml2011disease]. All terms under the “cancer” (DOID:162) parent term were selected and filtered for unspecific names of cancer (e.g. “neoplasm” or “carcinoma”). These cancer types were then matched with synonyms from the Unified Medical Language System (UMLS) Metathesaurus [@bodenreider2004unified] (2017AB), either through existing external reference links in the Disease Ontology or through exact string-matching on the main entity names. The additional synonyms in the UMLS were then added through this link. The genes list was built from the Entrez gene list and complemented with UMLS terms. Terms that overlapped with common words found in scientific literature (e.g. ice) were removed.

The drug list was curated from the WikiData resource [@vrandevcic2014wikidata]. All Wikidata entities that are drug instances (Wikidata identifier: Q12140) were selected using a SPARQL query. The generic name, brand name and synonyms were extracted where possible. This link was complemented by a custom list of general drug categories (e.g. chemotherapy, tyrosine kinase inhibitors, etc) and a list of inhibitors built using the previously discussed gene list. This allowed for the extraction of terms such as “EGFR inhibitors”. This was done because analysts are often interested in biomarkers associated with drug classes that target a specific gene, in addition to specific drugs.

All term lists were filtered with a stopwords list. This was based on the stopword list from the Natural Language Toolkit [@bird2006nltk] and the most frequent 5,000 words found in the Corpus of Contempory American English [@davies2009385] as well as custom set of terms. It was then merged with common words that occur as gene names (such as ICE).

@@

@@

The BioC corpora files were processed by the Kindred package. This NLP package used Stanford CoreNLP [@manning2014stanford] for processing in the original published version [@lever2017painless]. It was changed to Spacy [@honnibaljohnson] for the improved Python bindings in version 2 for this project. This provided easier integration and execution on a cluster without running a Java subprocess. Spacy was used for sentence splitting, tokenization and dependency parsing of the corpora files.

Exact string matching was then used against the tokenized sentences to extract mentions of cancer types, genes, drugs and variants. Longer terms were prioritised during extraction so that “non small cell lung cancer” would be extracted instead of just “lung cancer”. Variants were also extracted with a regular expression system for extracting protein coding point mutations (e.g. V600E).

Gene fusions (such as BCR-ABL1) were detected by identifying mentions of genes separated by a forward slash, hyphen or colon. If the two entities had no overlapping HUGO IDs, then it was flagged as a possible gene fusion and combined into a single entity. If there were overlapping IDs, it was deemed likely to be referring to the same gene. An example is HER2/neu which is frequently seen and refers to a single gene (ERBB2) and not a gene fusion.

Acronyms were also detected where possible by identifying terms in parentheses and checking the term before it, for instance “non-small cell lung carcinoma (NSCLC)”. This was done to remove entity mistakes where possible. The acronym detection method takes the short form (the term in brackets) and iterates backwards through the long form (the term before brackets) looking for potential matches for each letter. If the long form and short form has overlapping associated ontology IDs, they likely refer to the same thing and can be combined, as in the example above. If only one of the long form or short form has an associated ontology ID, they are combined and assigned the associated ontology ID. If both long form and short form have ontology IDs but there is no overlap, the short form is disregarded as the long form has more likelihood of getting the specific term correct.

Gene mentions that are likely associated with signalling pathways and not specific genes (e.g. “MTOR signalling”) are also removed using a simple pattern based on the words after the gene mention. One final post-processing step merges neighbouring terms with matching terms. So “HER2 neu” would be combined into one entity as the two terms (HER2 and neu) refer to the same gene.

@@

#keywords %>%
#  knitr::kable(
#    "html",
#    booktabs = TRUE, escape=FALSE,
#    caption = 'The five groups of filter strings  used to identify sentences that potentially discussed the four evidence types. Strings such as “sensitiv” are used to capture multiple words including “sensitive” and “sensitivity.'
#    ) %>%
#  kable_styling()

knitr::kable(
    keywords,
    booktabs = TRUE, escape=FALSE,
    caption = 'The five groups of search terms  used to identify sentences that potentially discussed the four evidence types. Strings such as “sensitiv” are used to capture multiple words including “sensitive” and “sensitivity".'
    )
```

With all biomedical documents parsed and entities tagged, all sentences were selected that mention at least one gene, at least one cancer and at least one variant. A drug was not required as only one (Predictive) of the four evidence types involves a drug entity. These sentences were enriched by filtering with certain keywords that are strongly associated with the different evidence items. The full list and groupings of keywords are shown in Table \@ref(tab:keywords). This grouping is done to make sure that each evidence type is represented reasonably equally in the training data. The General category with the keyword “marker” is included to catch additional sentences that discuss markers, which may relate to any of the four evidence types. Several of the keywords are stems in order to capture different forms of the word, e.g. prognosis or prognostic The acronym "DFS" which means "disease free survival" is also included as it was found in many sentences describing prognosis.

@@

A web platform for simple relation annotation was built using Bootstrap (https://getbootstrap.com/). This allowed annotators to work using a variety of devices, including their smartphones. The annotation system could be loaded with a set of sentences with entity annotations stored in a separate file (also known as standoff annotations). When provided with a relation pattern, for example “Gene/Cancer”, the system would search the input sentences and find all pairs of the given entity types in the same sentence. It would make sure that the two entities are not the same term, as in some sentences a token (or set of tokens) could be annotated as both a gene and a cancer, for instance “retinoblastoma”. For a sentence with 2 genes and 2 cancer types, it would find all four possible pairs of gene and cancer type.

@@

(ref:annotationplatform) A screenshot of the annotation platform that allowed expert annotators to select the relation types for different candidate relations in all of the sentences. The example sentence shown would be tagged using "Predictve/Prognostic" as it describes a prognostic marker.

@@

(ref:annotationOverview) An overview of the annotation process. Sentences are identified from the literature that describe cancers, genes, variants and optionally drugs and then filtered using search terms. The first test phase tried complex annotation of biomarker and variants together but was unsuccessful. The annotation task was split into two separate tasks for biomarkers and variants separately. Each task had a test phase and then the main phase on the 800 sentences that were used to create the gold set.

For the annotation stage (outlined in Fig \@ref(fig:annotationOverview)), an equal number of sentences were selected from each of the groups outlined in Table \@ref(tab:keywords). This guaranteed coverage of all four evidence types as the prognostic type dominated the other groups. If this step was not done, 100 randomly selected sentences would only contain 2 (on average) from the Diagnostic group. However, this sampling provided poor coverage of sentences that describe specific point mutations. Many precision oncology projects only focus on point mutations and so a further requirement was that 50% of sentences for annotation include a specific point mutation. All together, this sampling provides better coverage of the different omic events and evidence types that were of interest. Special care is required when evaluating models built on this customized training set as an unweighted evaluation would not be representative of the real literature.

Sentences that contain many permutations of relationships (e.g. a sentence with 5 genes and 5 cancer types mentioned) were removed. An upper limit of 5 possible relations was enforced for each sentence. This was done with the knowledge that the subsequent relation extraction step would have a greater false positive rate for sentences with very large number of possible relations. It was also done to make the annotation task more manageable. An annotation manual was constructed with examples of sentences that would and would not match the four evidence types. This was built in collaboration with CIViC curators. The annotation manual is available in our Github repository.

The annotation began with a test phase of 100 sentences that had poor annotator agreement and required a refinement of the annotation task outlined in this paragraph. The test phase allows the annotators to become accustomed to the annotation platform and make adjustments to the annotation manual to clarify misunderstandings. The first test phase (Biomarker + Variant) involved annotating sentences for ternary (gene, cancer, variant) or quaternary (gene, cancer, variant, drug) relationships. The ternary relationships included Diagnostic, Prognostic and Predisposing and the quaternary relationship was Predictive. A low F1-score inter-annotator agreement (average of 0.52) forced us to reconsider the annotation approach. This poor agreement was likely due to including variants within the annotations and provided a large combinatorial problem of exactly which entity mentions to include within a relationship. In order to simplify the problem, the task was split into two separate annotation tasks, the biomarker annotation and the variant annotation. The biomarker annotation involved binary (gene, cancer) and ternary (gene, cancer, drug) relations that described one of the evidence types. The Predictive and Prognostic evidence types were merged (as shown in Figure 2), to further reduce the annotation complexity. The Predictive/Prognostic annotations could be separated after tagging as relationships containing a drug would be Predictive and those without would be Prognostic. Any Prognostic relationship for a gene and cancer type that are in a Predictive relationship were removed. The variant annotation task (gene, variant) focused on whether a variant (e.g. deletion) was associated with a specific gene in the sentence.

With the redefined annotation task, six annotators were involved in biomarker annotation, all with knowledge of the CIViC platform and have experience interpreting patient cancer genome samples. Three annotators were involved in variant annotation, all with experience in cancer genomics. Both annotation tasks started with a new 100-sentence test phase to evaluate the redefined annotation tasks and resolve any ambiguity within the annotation manuals. Good inter-annotator agreement was achieved at this stage for both the biomarker annotation (average F1-score = 0.68) and variant annotation (average F1-score = 0.95). These 100 sentences were discarded as they exhibited a learning curve as annotators become comfortable with the task.

@@

### Application to PubMed and PMCOA

>>>>

## Abstract
Precision oncology involves analysis of individual cancer samples to understand the genes and pathways involved in the development and progression of a cancer. To improve patient care, knowledge of diagnostic, prognostic, predisposing and drug response markers is essential. Several knowledgebases have been created by different groups to collate evidence for these associations. These include the open-access Clinical Interpretation of Variants in Cancer (CIViC) knowledgebase. These databases rely on very time-consuming manual curation from highly-skilled experts who read and interpret the relevant biomedical literature. To aid in this curation and provide the greatest coverage for these databases, particularly CIViC, we propose the use of text mining approaches to extract these clinically relevant biomarkers from all available published literature. A group of cancer genomics experts annotated biomarkers and their clinical associations discussed in 800 sentences and achieved good inter-annotator agreement. We used a supervised learning approach to construct the CIViCmine knowledgebase (http://bionlp.bcgsc.ca/civicmine/) and extracted `r paper.sentenceCount` relevant sentences from PubMed abstracts and Pubmed Central Open Access full text papers. CIViCmine contains over `r paper.biomarkerCount` biomarkers associated with `r paper.geneCount` genes, `r paper.drugCount` drugs and `r paper.cancerCount` cancer types, representing `r paper.abstractCount` abstracts and `r paper.articleCount` full-text publications. Through integration with CIVIC, we provide a prioritised list of curatable biomarkers as well as a resource that is valuable to other knowledgebases and precision cancer analysts in general.

@@

The ability to stratify patients into groups that are clinically related is an important step towards a personalized approach to cancer. Over time, a growing number of biomarkers have been developed in order to select patients who are more likely to respond to certain treatments. These biomarkers have also been valuable for prognostic purposes and for understanding the underlying disease biology by defining different molecular subtypes of cancers that should be treated in different ways (e.g. *ERBB2*/*ESR1*/*PGR* testing in breast cancer [@onitilo2009breast]). Immunohistochemistry techniques are a primary approach for testing samples for diagnostic markers (e.g. CD15 and CD30 for Hodgkin’s disease [@rudiger1998differential]). Recently, the lower cost and increasing speed of sequencing has also allowed the DNA and RNA of individual patient samples to be characterized for clinical applications [@prasad2016precision]. Throughout the world, this technology is beginning to inform clinician decisions on which treatments to use [@shrager2014rapid]. Such efforts are dependent on comprehensive and current understanding of the clinical relevance of variants. For example, the Personalized Oncogenomics project at BC Cancer identifies somatic events in the genome such as point mutations, copy number variations and large structural changes and, in conjunction with gene expression data, generates a clinical report to provide an ‘omic picture of a patient’s tumor [@laskin2015lessons].

The high genomic variability observed in cancers means that each patient sample includes a large number of new mutations, many of which have never been documented before [@chang2016identifying]. The phenotypic impact of most of these mutations is difficult to discern. This problem is exacerbated by the driver/passenger mutation paradigm where only a fraction of mutations are essential to the cancer (drivers) while many others have occurred through mutational processes that are irrelevant to the cancer (passengers). An analyst trying to understand a new patient sample typically performs a literature review for each gene and specific variant. This is needed to understand its relevance in a cancer type, characterize the driver/passenger role of its observed mutations, and gauge the relevance for clinical decision making.

Several groups have built their own in-house knowledgebases, which are developed as analysts examine increasing numbers of cancer patient samples. This tedious and largely redundant effort represents a substantial interpretation bottleneck impeding the progress of precision medicine [@good2014organizing]. To encourage a collaborative effort, the CIViC knowledgebase (https://civicdb.org) was launched to provide a wiki-like, editable online resource where community contributed edits and additions are moderated by experts in order to maintain high quality [@griffith2017civic]. The resource provides information about clinically-relevant variants in cancer described in the peer-reviewed literature. Variants include protein-coding point mutations, copy number variations, epigenetic marks, gene fusions, aberrant expression levels and other ‘omic events. It supports four types of evidence associating biomarkers with different classes of clinical relevance (also known as evidence types).

Diagnostic evidence items describe variants that can help a clinician diagnose or exclude a cancer. For instance, the *JAK2* V617F mutation is a major diagnostic criterion for myeloproliferative neoplasms to identify polycythemia vera, essential thrombocythemia and primary myelofibrosis [@mesa2016myeloproliferative]. Predictive evidence items describe variants that help predict drug sensitivity or response and are valuable in deciding further treatments. Predictive evidence items often explain mechanisms of resistance in patients who progressed on a drug treatment. For example, the *ABL1* T315I missense mutation in the *BCR-ABL* fusion, predicts poor response to imatinib, a tyrosine kinase inhibitor that would otherwise effectively target *BCR-ABL*, in patients with chronic myeloid leukemia [@branford2003detection]. Predisposing evidence items describe germline variants that increase the likelihood of developing a particular cancer, such as *BRCA1* mutations for breast/ovarian cancer [@king2003breast] or *RB1* mutations for retinoblastoma [@harbour1998overview]. Lastly, prognostic evidence items describe variants that predict survival outcome. As an example, colorectal cancers that harbor a *KRAS* mutation are predicted to have worse survival [@phipps2013kras].

@@

Several other groups have created knowledgebases to aid clinical interpretation of cancer genomes. Many of these projects have joined the Variant Interpretation for Cancer Consortium (VICC, http://cancervariants.org/) to coordinate these efforts and have created a federated search mechanism to allow easier analysis across multiple knowledgebases [@wagner2018harmonized]. The CIViC project is co-leading this effort along with OncoKB [@chakravarty2017oncokb], the Cancer Genome Interpreter [@tamborero2018cancer], Precision Medicine Knowledge base [@huang2017cancer], Molecular Match, JAX-Clinical Knowledge base [@patterson2016clinical] and others.

@@

The literature on clinically relevant cancer mutations is growing at an extraordinary rate. For instance there were only 5 publications with BRAF V600E in title or abstract in PubMed in 2004 compared to 454 papers in 2017. In order to maintain a high quality and up-to-date knowledge-base, a curation pipeline must be established. This typically involves a queue for papers, triaging those that should be curated and then assignment to a highly experienced curator. This prioritisation step is important given the limited time of curators and the potentially vast number of papers to be reviewed. Prioritisation must identify papers that contain knowledge that is of current relevance to users of the knowledgebase. For instance, selecting papers for drugs that are no longer clinically approved would not be valuable to the knowledgebase.

Text mining methods have become a common approach to help prioritise literature curation. These methods fall broadly into two main categories, information retrieval (IR) and information extraction (IE). IR methods focus on paper-level information and can take multiple forms. Complex search queries for specific terms or paper metadata (helped by the MeSH term annotations of papers in biomedicine) are common tools for curators. More advanced document clustering and topic modelling systems can use semi-supervised methods to predict whether a paper would be relevant for curation. Examples of this approach include the document clustering method used for the ORegAnno project [@aerts2008text].

IE methods extract structured knowledge directly from the papers. This can take the form of entity recognition, by explicitly tagging mentions of biomedical concepts such as genes, drugs and diseases. A further step can involve relation extraction to understand the relationship discussed between tagged biomedical entities. This structured information can then be used to identify papers relevant for the knowledgebase. IE methods are also used for automated knowledgebase population without a manual curation step. For example, the miRTex knowledgebase, which collates microRNAs and their targets, uses automated relation extraction methods to populate the knowledgebase [@li2015mirtex]. Protein-protein interaction networks (such as STRING [@szklarczyk2016string]) are often built using automatically generated knowledgebases.

The main objective of this project was to identify frequently discussed cancer biomarkers that fit the CIViC evidence model but are not yet included in the CIViC knowledgebase. We developed an IE-based method to extract key parts of the evidence item: cancer type, gene, drug (where applicable) and the specific evidence type from published literature. This allows us to count the number of mentions of specific evidence items in abstracts and full text articles and compare against the CIViC knowledgebase. This paper will present our methods to develop this resource, known as CIViCmine (http://bionlp.bcgsc.ca/civicmine/). The main contributions of this work are an approach for knowledgebase construction that could be applied to many areas of biology and medicine, a machine learning method for extracting complicated relationships between four entity types, and extraction of relationships across the largest possible publically-accessible set of abstracts and full text articles. This resource, containing `r paper.biomarkerCount` gene-cancer associations with clinical relevance, is valuable to all cancer knowledgebases to aid their curation and also as a tool for precision cancer analysts searching for evidence supporting biomarkers not yet included in any other resource.

@@

The full PubMed and PubMed Central Open Access subset corpora was downloaded from the NCBI FTP website using the PubRunner infrastructure [@anekalla2017pubrunner]. These documents were converted to the BioC format for processing with the Kindred package [@lever2017painless]. HTML tags were stripped out and HTML special characters converted to Unicode. Metadata about the papers were retained including PubMed IDs, titles, journal information and publication date. Subsections of the paper were extracted using a customised set of acceptable section headers such as "Introduction", "Methods", "Results" and many synonyms of these  (accessible through the GitHub repository). The corpora were downloaded in bulk in order to not overload the EUtils RESTFUL service that is offered by the NCBI. In order to avoid duplications of publications in PMCOA and PubMed, the PMIDs of all documents included in PMCOA were used to filter out abstracts from the PubMed corpus. The updated files from PubMed were also processed to identify the latest version of each abstract to process.

@@

Term lists were curated for genes, diseases and drugs based on several resources. The cancer list was curated from a section of the Disease Ontology [@schriml2011disease]. All terms under the "cancer" (DOID:162) parent term were selected and filtered for nonspecific names of cancer (e.g. "neoplasm" or "carcinoma")  . These cancer types were then matched with synonyms from the Unified Medical Language System (UMLS) Metathesaurus [@bodenreider2004unified] (2017AB), either through existing external reference links in the Disease Ontology or through exact string-matching on the main entity names. The additional synonyms in the UMLS were then added through this link. The genes list was built from the Entrez gene list and complemented with UMLS terms. Terms that overlapped with common words found in scientific literature (e.g. ice) were removed.

The drug list was curated from the WikiData resource [@vrandevcic2014wikidata]. All Wikidata entities that are drug instances (Wikidata identifier: Q12140) were selected using a SPARQL query. The generic name, brand name and synonyms were extracted where possible. This list was complemented by a custom list of general drug categories (e.g. chemotherapy, tyrosine kinase inhibitors, etc) and a list of inhibitors built using the previously discussed gene list. This allowed for the extraction of terms such as "EGFR inhibitors". This was done because analysts are often interested in and publications often discuss biomarkers associated with drug classes that target a specific gene, in addition to specific drugs.

All term lists were filtered with a stopwords list. This was based on the stopword list from the Natural Language Toolkit [@bird2006nltk] and the most frequent 5,000 words found in the Corpus of Contempory American English [@davies2009385] as well as a custom set of terms  . It was then merged with common words that occur as gene names (such as ICE).

@@

The word lists and tools used to generate them are accessible through the BioWordlists project (https://github.com/jakelever/biowordlists) and data can be found in the Zenodo repository ((http://doi.org/10.5281/zenodo.1286661).

@@

The BioC corpora files were processed by the Kindred package. This NLP package used Stanford CoreNLP [@manning2014stanford] for processing in the original published version [@lever2017painless]. For this work, it was changed to Spacy [@honnibaljohnson] for the improved Python bindings in version 2 for this project.   This provided easier integration and execution on a cluster without running a Java subprocess. Spacy was used for sentence splitting, tokenization and dependency parsing of the corpora files.

Exact string matching was then used against the tokenized sentences to extract mentions of cancer types, genes, drugs and variants. Longer terms were prioritised during extraction so that "non small cell lung cancer" would be extracted instead of just "lung cancer". Variants were also extracted with a regular expression system for extracting protein coding point mutations (e.g. V600E).

Gene fusions (such as *BCR-ABL1*) were detected by identifying mentions of genes separated by a forward slash, hyphen or colon. If the two entities had no overlapping HUGO IDs, then it was flagged as a possible gene fusion and combined into a single entity. If there were overlapping IDs, it was deemed likely to be referring to the same gene. An example is *HER2/neu* which is frequently seen and refers to a single gene (*ERBB2*) and not a gene fusion.

Acronyms were also detected, where possible, by identifying terms in parentheses and checking the term before it, for instance "non-small cell lung carcinoma (NSCLC)". This was done to remove entity mistakes where possible. The acronym detection method takes the short form (the term in brackets) and iterates backwards through the long form (the term before brackets) looking for potential matches for each letter. If the long form and short form has overlapping associated ontology IDs, they likely refer to the same thing and can be combined, as in the example above. If only one of the long form or short form has an associated ontology ID, they are combined and assigned the associated ontology ID. If both long form and short form have ontology IDs but there is no overlap, the short form is disregarded as the long form has more likelihood of getting the specific term correct  .

Gene mentions that are likely associated with signalling pathways and not specific genes (e.g. "MTOR signalling") are also removed using a simple pattern based on the words after the gene mention. One final post-processing step merges neighbouring terms with matching terms. So "HER2 neu" would be combined into one entity as the two terms (*HER2* and *neu*) refer to the same gene.

@@

#keywords %>%
#  knitr::kable(
#    "html",
#    booktabs = TRUE, escape=FALSE,
#    caption = 'The five groups of filter strings  used to identify sentences that potentially discussed the four evidence types. Strings such as "sensitiv" are used to capture multiple words including "sensitive" and "sensitivity".'
#    ) %>%
#  kable_styling()

knitr::kable(
    keywords,
    booktabs = TRUE, escape=FALSE,
    caption = 'The five groups of search terms used to identify sentences that potentially discussed the four evidence types  . Strings such as "sensitiv" are used to capture multiple words including "sensitive" and "sensitivity".'
    )
```

With all biomedical documents parsed and entities tagged, all sentences were selected that mention at least one gene, at least one cancer and at least one variant. A drug was not required as only one (predictive) of the four evidence types involves a drug entity. These sentences were enriched by filtering with certain keywords that are strongly associated with the different evidence items. The full list and groupings of keywords are shown in Table \@ref(tab:keywords)). This grouping is done to make sure that each evidence type is represented reasonably equally in the training data. The General category with the keyword "marker" is included to catch additional sentences that discuss markers, which may relate to any of the four evidence types. Several of the keywords are stems in order to capture different forms of the word, e.g. prognosis or prognostic. The acronym "DFS" which means "disease free survival"  is also included as it was found in many sentences describing prognosis.

@@

A web platform for simple relation annotation was built using Bootstrap (https://getbootstrap.com/). This allowed annotators to work using a variety of devices, including their smartphones. The annotation system could be loaded with a set of sentences with entity annotations stored in a separate file (also known as standoff annotations). When provided with a relation pattern, for example "Gene/Cancer", the system would search the input sentences and find all pairs of the given entity types in the same sentence. It would make sure that the two entities are not the same term, as in some sentences a token (or set of tokens) could be annotated as both a gene name and a cancer type (e.g. "retinoblastoma"). For a sentence with 2 genes and 2 cancer types, it would find all four possible pairs of gene and cancer type.

@@

(ref:annotationplatform) A screenshot of the annotation platform that allowed expert annotators to select the relation types for different candidate relations in all of the sentences. The example sentence shown describes a prognostic marker.

@@

(ref:annotationOverview) An overview of the annotation process. Sentences are identified from the literature that describe cancers, genes, variants and optionally drugs before being filtered using search terms. The first test phase tried complex annotation of biomarker and variants together but was unsuccessful. The annotation task was split into two separate tasks for biomarkers and variants separately. Each task had a test phase and then the main phase on the 800 sentences that were used to create the gold set.

For the annotation step (outlined in Fig \@ref(fig:annotationOverview)), the annotated data set (known as the gold set) was constructed using a consensus of multiple annotators. An equal number of sentences were selected from each of the groups outlined in Table \@ref(tab:keywords). This guaranteed coverage of all four evidence types as otherwise the prognostic type dominated the other groups. If this step was not done, 100 randomly selected sentences would only contain 2 (on average) from the diagnostic group. However, this sampling provided poor coverage of sentences that describe specific point mutations. Many precision oncology projects only focus on point mutations and so a further requirement was that 50% of sentences for annotation include a specific point mutation. All together, this sampling provides better coverage of the different omic events and evidence types that were of interest. Special care is required when evaluating models built on this customized training set as an unweighted evaluation would not be representative of the real literature.

Sentences that contain many permutations of relationships (e.g. a sentence with 6 genes and 4  cancer types mentioned) were removed. An upper limit of 5 possible relations was enforced for each sentence. This was done with the knowledge that the subsequent relation extraction step would have a greater false positive rate for sentences with very large number of possible relations. It was also done to make the annotation task more manageable. An annotation manual was constructed with examples of sentences that would and would not match the four evidence types. This was built in collaboration with CIViC curators and is available in our Github repository (https://github.com/jakelever/civicmine).

Each annotation task began with a test phase of 100 sentences. This allows the annotators to become accustomed to the annotation platform and make adjustments to the annotation manual to clarify misunderstandings. The first test phase (Biomarker + Variant) involved annotating sentences for ternary (gene, cancer, variant) or quaternary (gene, cancer, variant, drug) relationships. The ternary relationships included diagnostic, prognostic and predisposing and the quaternary relationship was predictive. A low F1-score inter-annotator agreement (average of 0.52) forced us to reconsider the annotation approach. From our evaluation, we determined that this poor agreement was likely due to including variants within the annotations. This created a large combinatorial problem where it was often unclear exactly which entity mentions to include within a relationship. In order to simplify the problem, the task was split into two separate annotation tasks, the biomarker annotation and the variant annotation. The biomarker annotation involved binary (gene, cancer) and ternary (gene, cancer, drug) relations that described one of the evidence types. The predictive and prognostic evidence types were merged (as shown in Figure 2), to further reduce the annotation complexity. The predictive/prognostic annotations could be separated after tagging as relationships containing a drug would be predictive and those without would be prognostic.   A further postprocessing step to generate the gold set involved identifying prognostic relationships that overlapped with predictive relationships (i.e. shared the same gene and cancer type in a sentence) and removing them. The variant annotation task (gene, variant) focused on whether a variant (e.g. deletion) was associated with a specific gene in the sentence.

With the redefined annotation task, six annotators were involved in biomarker annotation, all with knowledge of the CIViC platform and having experience interpreting patient cancer variants in a clinical context. Three annotators (one of whom was involed in the biomarker annotation)   were involved in variant annotationand they all had experience in cancer genomics. Both annotation tasks started with a new 100-sentence test phase to evaluate the redefined annotation tasks and resolve any ambiguity within the annotation manuals. Good inter-annotator agreement was achieved at this stage for both the biomarker annotation (average F1-score = 0.68) and variant annotation (average F1-score = 0.95). These 100 sentences were discarded as they exhibited a learning curve as annotators become comfortable with the task.

@@

### Application to PubMed and PMCOA with Updates
